---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Atopic Dermatitis (Non-lesional skin) Case Study 

In this chapter, we evaluate the contributors of agreement between CellChat and CellPhoneDB inference results, using the publicly available scRNA-seq dataset of human atopic dermatitis skin samples from the study by [He et. al (2020)](https://www.jacionline.org/article/S0091-6749(20)30182-2/fulltext). The processed dataset provided by the developers of CellChat can be downloaded via Figshare [here](https://doi.org/10.6084/m9.figshare.13520015). We downloaded the dataset and performed further processing, and subset the data into the non-lesional and lesional skin samples, which we saved as `seu.NL` and `seu.LS` respectively. The code for processing the data is available in our GitHub repository [here](https://github.com/nigelhojinker/scpeakeR/blob/main/data-raw/demo_data.md).

In this section, we document the process used to compare the inference results generated by CellChat and CellPhoneDB using `seu.NL` as the first example. We compiled functions in the `deploy.R` script, which you may refer to [here](https://github.com/nigelhojinker/MCC_regression/blob/main/R/deploy.R). 

## Load Data and Packages

```{r}
pacman::p_load(scpeakeR, ggthemes, ComplexHeatmap, circlize)
setwd(this.path::here())
rm(list = ls())

use_condaenv("scpeaker")

data(seu.NL, step1_map, CellChatDB.human)
load("data/CellPhoneDB_ori_human.rda")    # CellChat-compatible CellPhoneDB resource in .rda format
```

## Load required functions

```{r}
source("R/deploy.R")
```

## Create result directory if it doesn't exist
```{r, eval = FALSE}
sapply("results", dir.create, recursive = TRUE)
```

## seu.NL data summary

`seu.NL` is a Seurat object containing 10,353 genes across 2,233 cells. 

```{r}
seu.NL
```

### seu.NL metadata

```{r}
seu.NL@meta.data %>% head()
```

### seu.NL cell type clusters

```{r}
seu.NL@meta.data %>% tabyl(labels)
```

# Conduct CellChat and CellPhoneDB analysis across all configurations

## Create table of all configurations to test

```{r}
grid <- expand.grid(method   = c("cellphonedb", "cellchat"),
                    mean     = c("t10", "t25", "triMean"),
                    database = c("CPDB", "CCDB"))

grid %>% knitr::kable()
```

## Parse mean type and threshold from grid

```{r}
parse_mean <- function(x){
  if (x == "t10") {
    list(type = "thresholdedMean", threshold = 0.1)
  } else if (x == "t25"){
    list(type = "thresholdedMean", threshold = 0.25)
  } else if (x == "triMean"){
    list(type = "triMean", threshold = 0)
  }
}
```

## Run analysis across all configurations

```{r, eval = FALSE}
step1_perm_NL <- apply(grid, 1, function(row){
  parsed <- parse_mean(row["mean"])

  cat("Running analysis with method:", row["method"], "\n",
      "mean:", parsed$type, "\n",
      "threshold:", parsed$threshold, "\n",
      "database:", row["database"], "\n")

  if (row["method"] == "cellchat") {
    run_cellchatv2(seu.NL,    mean = parsed$type, database = row["database"], threshold = parsed$threshold)
  } else if(row["method"] == "cellphonedb") {
    run_cellphonedbv2(seu.NL, mean = parsed$type, database = row["database"], threshold = parsed$threshold)
  }
})

names(step1_perm_NL) <- apply(grid, 1, paste, collapse = "_")

save(step1_perm_NL, file = "results/step1_perm_NL.rda")
```

# Compute the MCC for each pairwise comparison of the 12 configurations

Now that we have generated the inference results across all 12 configurations, we can compute the MCC for each pairwise comparison of the configurations. This will allow us to evaluate the agreement between the different methods, mean types and databases, and identify the factors contributing to agreement between CellChat and CellPhoneDB.

```{r, echo = FALSE}
load("results/step1_perm_NL.rda")
```

```{r, eval = FALSE}
seuNL_corr_s1 <- compute_corr(step1_perm_NL)

save(seuNL_corr_s1, file = "results/seuNL_corr_s1.rda") 
```

```{r, echo = FALSE}
load("results/seuNL_corr_s1.rda") 
```

```{r}
seuNL_corr_s1 %>% dim()

seuNL_corr_s1 %>% 
  head() %>%
  knitr::kable()
```

# Visualize the correlation matrix as a heatmap

```{r}
#| fig-width: 10
#| fig-height: 8
#| out-width: 100%

create_heatmap(seuNL_corr_s1, step1_perm_NL, "mcc")
```

# Multiple linear regression analysis

Here, we perform a multiple linear regression analysis to evaluate the effect size of each factor (method, mean type and database) on the agreement between CellChat and CellPhoneDB, as measured by the MCC. We will fit a linear model with the MCC as the response variable and the method, mean type and database as predictor variables. This will allow us to quantify the contribution of each factor to the agreement between the two tools.

## Create a data frame for regression analysis

We removed the diagonal entries of the correlation matrix, as they correspond to self-self comparisons which is always 1 and does not provide meaningful information for the regression analysis. We also removed the upper triangle of the matrix to avoid duplicate comparisons, as the correlation matrix is symmetric.

We added columns `Method`, `Mean_Type` and `Database` to indicate if the pairwise comparisons were conducted between two configurations with the same method, mean type or database respectively. For example, if the pairwise comparison was conducted between `cellchat_t10_CCDB` and `cellphonedb_t10_CCDB`, the `Method` column would be Diff (different methods), the `Mean_Type` column would be Same (same mean type) and the `Database` column would be Same (same database).

```{r}
regression_data <- format_data(seuNL_corr_s1)

regression_data %>% head() %>% knitr::kable()
```

## Fit a linear model

```{r}
fit <- lm(mcc ~ Method + Mean_Type + Database, data = regression_data)

summary(fit)

broom::tidy(fit) %>% knitr::kable()
```

# Discussion

The fitted regression model was:

$$
\begin{aligned}
MCC =\;& 0.7769\, \\ 
      & - 0.0762\, I(\text{Method}) \\
      & - 0.1148\,   I(\text{Mean\_Type}) \\
      & - 0.4922\,   I(\text{Database})
\end{aligned}
$$

where $I = 0$ indicates that the factor was the same between the analyses, and $I = 1$ indicates that the factor differed.

The overall regression was statistically significant, explaining a substantial proportion of the variance in the computed MCC values (R-squared = 0.8203). Under the reference condition, where method, mean type and database were identical, the expected MCC was 0.7769. Holding the other factors constant, MCC decreased by 0.0762 when the method differed, by 0.1148 when the mean type differed, and by 0.4922 when the database differed. All predictors displayed a significant decrease (p < 0.05) in the MCC value when the variable differed, with database choice exerting the largest effect, followed by mean type and method. 

These results indicate that across CCC analyses performed using CellChat and CellPhoneDB, most of the variability in agreement is attributed to database choice. In contrast, changes in inference method or mean type used to compute average gene expression per cell type cluster contributed comparatively smaller shifts in MCC when the database used is held constant. Together, these findings suggest that ligand-receptor prior knowledge is the most important factor among the three predictors tested to understand agreement between CellChat and CellPhoneDB inference results.

